---
title: "U.S. Airline Networks and Their Effects on Prices"
author: "Grant Hoffman"
date: "`r Sys.Date()`"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
pacman::p_load(tidyverse,skimr,stargazer,pander,kableExtra, ggplot2, haven, broom, sandwich, lmtest, memisc, xtable, AER, plm, ggrepel, maps)
options(knitr.table.format = "latex")

theme_set(theme_bw())
```

## Introduction 
- What Effects Flight Fares?
  - Economies of Scale
    - Bigger Planes 
    - Larger Airports
  - Route Alternatives
    - Number of routes offered from point A to B
    - Competing airlines on the route
    
- Effect of Airline Mergers
  - Better economies of scale
  - Less Route alternatives and competition

## 2007 Flights
```{r, echo = FALSE, include = FALSE}
## Load in Data
Loc_DO <- read_csv("Data/Airport_DO_Locations.csv")

Loc_Lay <- read_csv("Data/Airport_Lay_Locations.csv")
Loc <- Loc_DO %>% 
  full_join(y = Loc_Lay)
Loc <- Loc %>% 
  mutate(Layover = as.character(Layover),
         Lay_lat = as.numeric(Lay_lat),
         Lay_long = as.numeric(Lay_long))

## Create the map from 2007 flights
RNGversion("4.0.0")
set.seed(3457)
Loc <- Loc[sample(1:nrow(Loc)), ] 
Loc_07 <- Loc %>% 
  filter(Year == 2007,
         Quarter == 1,
         Dest_long > -130,
         Org_long > -130,
         Dest_lat < 60,
         Dest_lat > 25,
         Org_lat < 60,
         Org_lat > 25)
Loc_07 <-  Loc_07[1:100,]
Loc_07D <- Loc_07[1:100,] %>% 
  filter(is.na(Layover))
Loc_07L <- Loc_07[1:100,] %>% 
  filter(!is.na(Layover))

Loc_11 <- Loc %>% 
  filter(Year == 2011,
         Quarter == 4,
         Dest_long > -130,
         Org_long > -130,
         Dest_lat < 60,
         Dest_lat > 25,
         Org_lat < 60,
         Org_lat > 25)
Loc_11 <-  Loc_11[1:100,]
Loc_11D <- Loc_11[1:100,] %>% 
  filter(is.na(Layover))
Loc_11L <- Loc_11[1:100,] %>% 
  filter(!is.na(Layover))
```

```{r, Hub and Spoke Display, fig.cap = "Sample of 100 Flights from Quarter 1 of 2007"}
usMap <- borders("state",color = "grey",fill = "white")
ggplot(mapping = aes(color = OpCarrier)) + usMap +
  geom_curve(data=Loc_07D,
             aes(x=Org_long, y=Org_lat, xend=Dest_long, yend=Dest_lat),
             # color ="green",
             size=.5,
             curvature = 0) +
  geom_curve(data = Loc_07L,
             aes(x=Org_long, y = Org_lat, xend = Lay_long, yend = Lay_lat),
             #color = "blue",
             size = 0.5,
             curvature = 0.3) +
  geom_curve(data = Loc_07L,
             aes(x=Lay_long, y = Lay_lat, xend = Dest_long, yend = Dest_lat),
             #color = "blue",
             size = 0.5,
             curvature = 0.3) +
  geom_point(data=Loc_07,
             aes(x=Org_long, y=Org_lat), 
             color ="blue",
             size=1.5,
             alpha = 0.5) +
  geom_point(data = Loc_07L,
             aes(x = Lay_long, y = Lay_lat),
             color = "red") +
  geom_point(data=Loc_07,
             aes(x=Dest_long, y=Dest_lat), 
             color ="blue",
             alpha = 0.5) +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.ticks=element_blank(),
        plot.title=element_text(hjust=0.5, size=12)) +
  ggtitle("Sample of 100 Flights from Quarter 1 of 2007")
```

## 2011 Flights
```{r, Hub and Spoke 2011, fig.cap = "Sample of 100 Flights from Quarter 4 of 2011",}
usMap <- borders("state",color = "grey",fill = "white")
ggplot(mapping = aes(color = OpCarrier)) + usMap +
  geom_curve(data=Loc_11D,
             aes(x=Org_long, y=Org_lat, xend=Dest_long, yend=Dest_lat),
             #color ="green",
             size=.5,
             curvature = 0) +
  geom_curve(data = Loc_11L,
             aes(x=Org_long, y = Org_lat, xend = Lay_long, yend = Lay_lat),
             #color = "blue",
             size = 0.5,
             curvature = 0.3) +
  geom_curve(data = Loc_11L,
             aes(x=Lay_long, y = Lay_lat, xend = Dest_long, yend = Dest_lat),
             #color = "blue",
             size = 0.5,
             curvature = 0.3) +
  geom_point(data=Loc_11,
             aes(x=Org_long, y=Org_lat), 
             color ="blue",
             size=1.5,
             alpha = 0.5) +
  geom_point(data = Loc_11L,
             aes(x = Lay_long, y = Lay_lat),
             color = "red") +
  geom_point(data=Loc_11,
             aes(x=Dest_long, y=Dest_lat), 
             color ="blue",
             alpha = 0.5) +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.ticks=element_blank(),
        plot.title=element_text(hjust=0.5, size=12))
```

## Flight Tables

```{r Flight Table, fig.cap = "Amount of Flights from most popular Airports"}
Loc_fac <- Loc %>% filter((Year == 2007 & Quarter == 1) | (Year == 2011 & Quarter == 4)) %>% 
  dplyr::select(Year,Quarter,Origin,Destination,Layover) %>% 
  pivot_longer(cols = c(Destination,Origin,Layover), names_to = "Included Airports", values_to = "Airports") 

Loc_fac %>% group_by(Airports) %>%  
  filter(n() > 1250 & !is.na(Airports)) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = Airports, fill = factor(Year))) +
  geom_bar(position = "Dodge") +
  labs(fill = "Year", y = "Number of Flights") +
  theme(legend.position = "bottom")
```

## Brueckner, Dyer, Spliller (1992)
- More spoke traffic means lower fares
- Competition does not lower fares across the board
- Higher traffic in the network lowers fares

## Theoretical Framework from Brueckner, Dyer, and Spiller
```{r, Hub and Spoke Figure,fig.align = "center", fig.height=2.5, fig.width=2.5}
xcoord <- c(0,0,-0.5,0.5)
ycoord <- c(0,0.5,-0.5,-0.5)
names <- c("H","A","B","C")

Hub_spoke <- data.frame(xcoord,ycoord,names)

Hub_spoke_plot <- Hub_spoke %>% ggplot() + 
  geom_point(mapping = aes(x = xcoord, y = ycoord)) +
  geom_segment(mapping = aes(x = xcoord[1], y = ycoord[1], xend = xcoord[2], yend = ycoord[2])) +
  geom_segment(mapping = aes(x = xcoord[1], y = ycoord[1], xend = xcoord[3], yend = ycoord[3])) +
  geom_segment(mapping = aes(x = xcoord[1], y = ycoord[1], xend = xcoord[4], yend = ycoord[4])) +
  geom_text(mapping = aes(x = xcoord, y = ycoord, label = names), hjust = c(-0.75,-0.5,0,0), vjust = c(0,0,-1,0)) +
  theme_void()

Hub_spoke_plot
```

<!-- ## Theoretical Framework from Brueckner, Dyer, and Spiller Continued -->
<!-- ```{r, Hub and Spoke Figure 2,fig.align = "center", fig.height=2.5, fig.width=2.5} -->
<!-- xcoord2 <- c(-1,1,-2,2,0,0) -->
<!-- ycoord2 <- c(0,0,0,0,-0.2,0.2) -->
<!-- names2 <- c("A","B","E","F","C","D") -->

<!-- Hub_spoke2 <- data.frame(xcoord2,ycoord2,names2) -->

<!-- Hub_spoke_plot2 <- Hub_spoke2 %>% ggplot() +  -->
<!--   geom_point(mapping = aes(x = xcoord2, y = ycoord2)) + -->
<!--   geom_curve(mapping = aes(x = xcoord2[1], y = ycoord2[1], xend = xcoord2[2], yend = ycoord2[2], -->
<!--                linetype = "Carrier 1"), -->
<!--              curvature = 0.2) + -->
<!--   geom_segment(mapping = aes(x = xcoord2[1], y = ycoord2[1], xend = xcoord2[2], yend = ycoord2[2], -->
<!--                              linetype = "Carrier 2")) + -->
<!--   geom_segment(mapping = aes(x = xcoord2[1], y = ycoord2[1], xend = xcoord2[3], yend = ycoord2[3], -->
<!--                linetype = "Carrier 1")) + -->
<!--   geom_segment(mapping = aes(x = xcoord2[2], y = ycoord2[2], xend = xcoord2[4], yend = ycoord2[4], -->
<!--                              linetype = "Carrier 2")) + -->
<!--   geom_segment(mapping = aes(x = xcoord2[1], y = ycoord2[1], xend = xcoord2[5], yend = ycoord2[5], -->
<!--                linetype = "Carrier 1")) + -->
<!--   geom_segment(mapping = aes(x = xcoord2[1], y = ycoord2[1], xend = xcoord2[6], yend = ycoord2[6], -->
<!--                linetype = "Carrier 1")) + -->
<!--   geom_segment(mapping = aes(x = xcoord2[2], y = ycoord2[2], xend = xcoord2[5], yend = ycoord2[5], -->
<!--                              linetype = "Carrier 2")) + -->
<!--   geom_segment(mapping = aes(x = xcoord2[2], y = ycoord2[2], xend = xcoord2[6], yend = ycoord2[6], -->
<!--                              linetype = "Carrier 2")) + -->
<!--   geom_text(mapping = aes(x = xcoord2, y = ycoord2, label = names2), vjust = c(-1,-1,-1,-1,2,3), size = 2) + -->
<!--   theme_void() + -->
<!--   # theme(text = element_text(size = 50)) + -->
<!--   scale_linetype_manual(values = c("dashed","solid")) + -->
<!--   labs(linetype = 'Carrier') + -->
<!--   theme(legend.key.size = unit(1, 'cm'), -->
<!--         legend.text = element_text(size = 4)) -->
<!-- Hub_spoke_plot2 -->
<!-- #ggsave(filename = "Figures/Hub and Spoke Model2.png", Hub_spoke_plot2)  -->

<!-- ``` -->

<!-- ## Theoretical Framework from Brueckner, Dyer, and Spiller Continued -->
<!-- ```{r, Hub and Spoke Figure 3,fig.align = "center", fig.height=2.5, fig.width=2.5} -->

<!-- carrier <- c("Carrier 1", "Carrier 2") -->
<!-- Hub_spoke_plot3 <- Hub_spoke %>% ggplot() +  -->
<!--   geom_point(mapping = aes(x = xcoord, y = ycoord)) + -->
<!--   geom_segment(mapping = aes(x = xcoord[1], y = ycoord[1], xend = xcoord[2], yend = ycoord[2], -->
<!--                linetype = "Major Carrier", -->
<!--                color = "Major Carrier")) + -->
<!--   geom_segment(mapping = aes(x = xcoord[1], y = ycoord[1], xend = xcoord[3], yend = ycoord[3], -->
<!--                linetype = "Major Carrier", -->
<!--                color = "Major Carrier")) + -->
<!--   geom_segment(mapping = aes(x = xcoord[1], y = ycoord[1], xend = xcoord[4], yend = ycoord[4], -->
<!--                linetype = "Major Carrier", -->
<!--                color = "Major Carrier")) + -->
<!--   geom_segment(mapping = aes(x = xcoord[2], y = ycoord[2], xend = xcoord[3], yend = ycoord[3], -->
<!--                linetype = "Low Cost Carrier", -->
<!--                color = "Low Cost Carrier"))+ -->
<!--   geom_text(mapping = aes(x = xcoord, y = ycoord, label = names), hjust = c(-0.75,-0.5,0,0), vjust = c(0,0,-1,0)) + -->
<!--   theme_void() + -->
<!--   scale_color_manual(values = c("red","black")) + -->
<!--   scale_linetype_manual(values = c("dashed","solid")) + -->
<!--   labs(linetype = 'Carrier', -->
<!--        color = 'Carrier') + -->
<!--   theme(legend.key.size = unit(1, 'cm'), -->
<!--         legend.text = element_text(size = 4)) -->
<!-- Hub_spoke_plot3 -->
<!-- #ggsave(filename = "Figures/Hub and Spoke Model3.png", Hub_spoke_plot3)  -->
<!-- ``` -->

<!-- ## Theoretical Framework from Brueckner, Dyer, and Spiller Continued -->
<!-- ```{r, Hub and Spoke Figure 4,fig.align = "center", fig.height=2.5, fig.width=2.5} -->

<!-- Hub_spoke_plot <- Hub_spoke %>% ggplot() +  -->
<!--   geom_point(mapping = aes(x = xcoord, y = ycoord)) + -->
<!--   geom_segment(mapping = aes(x = xcoord[1], y = ycoord[1], xend = xcoord[2], yend = ycoord[2], color = "B-A"), -->
<!--                arrow = arrow(length = unit(0.25, "inches"))) + -->
<!--   geom_segment(mapping = aes(x = xcoord[3], y = ycoord[3], xend = xcoord[1], yend = ycoord[1], color = "B-A"), -->
<!--                arrow = arrow(length = unit(0.25, "inches"))) + -->
<!--   geom_curve(mapping = aes(x = xcoord[1], y = ycoord[1], xend = xcoord[4], yend = ycoord[4], color = "B-C"), -->
<!--              curvature = -0.2, -->
<!--              arrow = arrow(length = unit(0.25, "inches")), -->
<!--              linetype = "dashed") + -->
<!--   geom_curve(mapping = aes(x= xcoord[3], y = ycoord[3], xend = xcoord[1], yend = ycoord[1], color = "B-C"), -->
<!--              curvature = -0.2, -->
<!--              arrow = arrow(length = unit(0.25, "inches")), -->
<!--              linetype ="dashed") + -->
<!--   geom_curve(mapping = aes(x= xcoord[3], y = ycoord[3], xend = xcoord[1], yend = ycoord[1], color = "B-H"), -->
<!--              curvature = 0.2, -->
<!--              arrow = arrow(length = unit(0.25, "inches")), -->
<!--              linetype = "dashed") + -->
<!--   geom_curve(mapping = aes(x = xcoord[4], y = ycoord[4], xend = xcoord[1], yend = ycoord[1], color = "C-A"), -->
<!--              curvature = -0.2, -->
<!--              arrow = arrow(length = unit(0.25, "inches")), -->
<!--              linetype = "dashed") + -->
<!--   geom_curve(mapping = aes(x = xcoord[1], y = ycoord[1], xend = xcoord[2], yend = ycoord[2], color = "C-A"), -->
<!--              curvature = -0.2, -->
<!--              arrow = arrow(length = unit(0.25, "inches")), -->
<!--              linetype = "dashed") + -->
<!--   geom_curve(mapping = aes(x = xcoord[1], y = ycoord[1], xend = xcoord[2], yend = ycoord[2], color = "H-A"), -->
<!--              curvature = 0.2, -->
<!--              arrow = arrow(length = unit(0.25, "inches")), -->
<!--              linetype = "dashed") + -->
<!--   geom_text(mapping = aes(x = xcoord, y = ycoord, label = names), hjust = c(-0.75,-0.5,0,0), vjust = c(0,0,-1,0)) + -->
<!--   theme_void() + -->
<!--   labs(color = 'Trip') + -->
<!--   theme(legend.key.size = unit(1, 'cm'), -->
<!--         legend.text = element_text(size = 4)) -->
<!-- #ggsave(filename = "Figures/Hub_Spoke_Example.png", Hub_spoke_plot) -->
<!-- Hub_spoke_plot -->
<!-- ``` -->

- This is a hub and spoke network of airports with H as the hub
- $D(Q_{ij})$ is the inverse demand function where $Q_{ij}$ is the traffic for a round trip from i to j

## Theoretical Framework Continued
- $R(Q_{ij}) = Q_{ij}*D(Q_{ij})$ is revenue for a round trip from i to j
- Total Revenue $R(Q)$ for the above hub and spoke is then: 
\begin{equation}
R(Q_{AB}) + R(Q_{AC}) + R(A_{BC}) + R(Q_{AH}) + R(Q_{BH}) + R(Q_{CH})
\end{equation}
- $c(Q_{ij})$ is the cost of a round trip from i to j
- Since there are increasing returns to scale $c' > 0$ and $c'' < 0$ 
- Profit is maximized by setting marginal revenue $R'(Q)$ equal to marginal cost $c'(Q)$

## Theoretical Framework Continued
- In a hub and spoke network there is no direct travel from spokes A to B or A to C or B to C
- Profit maximization for a spoke node A to the hub node H is simply: 
\begin{equation} 
R'(Q_{AH}) = c'(Q_{AB} + Q_{AC} + Q_{AH})
\end{equation}
- Profit maximization for a spoke node A to another spoke B is more complicated:
- Because there are no direct flights then marginal revenue for a flight from A to B would be:
\begin{equation}
R'(Q_{AB}) = R'(Q_{AH}) + R'(Q_{BH}) 
\end{equation}
- Substituting in the marginal cost functions yeilds:
\begin{equation}
R'(Q_{AB}) = c'(Q_{AB} + Q_{AC} + Q_{AH} + Q_{BA} + Q_{BC} + Q_{BH}) 
\end{equation}

## Applying the framework
- As the Airlines increasingly convert to hub and spoke models what is the effect on prices?
  - It depends on how strong the economies of density are.
  - Hub and spoke models increase the ability for airlines to use bigger planes and increase scaling
  - Hub and spoke models also require more connections which increases the number of flights.
  - which prevails? 
  
## Huschelrath and Muller (2014)
- Analyzes the effect of America West - US Airlines merger in 2005
- Uses a differences-in-differences appraoch fixing effects across route and ex-ante carrier
- Two types of routes used
  - Direct routes
  - Single connections
- Compares the price effects on each route depending on the type of competition between the airlines prior to the merger
- Found that on routes where competition between airlines took place there was an increase in fares but on routes that saw almost no competition between airlines there was a decrease in prices that was larger
- Overall increase in consumer welfare

## My paper
- My paper will combine the two approaches
- Using the network characteristics of the airports along with the route competition to enhance the analyses done by Huschelrath and Muller
- The goal is to use a differences-in-differences approach and control for network characteristics to analyze the consumer welfare effects of the Northwest-Delta Merger of 2010. 


